# Registration analysis

<style type="text/css">
  body{
  font-family: Futura, Helvetica, Arial;
}
</style>

<br>

```{r setup, include = FALSE, echo = FALSE}

library(here)
source(here("R", "01_startup.R"))

knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(include = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)

# Function for checking file modifications
mtime <- function(files) lapply(Sys.glob(files), function(x) file.info(x)$mtime)

# Set this to TRUE to output PDF figures
build_figures <- TRUE
# regular <- NULL

qload(here("output", "/geometry.qsm"))
qload(here("output", "/str_processed.qsm"))
STA <- qread(here("output", "/STA.qs"))
home_sales <- qread("output/home_sales.qs")


```
``` {r objects, include = TRUE}

start_reg <- as.Date("2019-11-04")
end_reg <- as.Date("2020-09-04")
start_mora <- as.Date("2020-09-05")
end_mora <- as.Date("2021-06-22")

# Get properties active during the period when the registration was into effect
reg_property <- 
  property %>% 
  filter(scraped >= start_reg) %>% 
  st_transform(32618)

n_reg_property <- 
  reg_property %>% 
  nrow() %>% 
  prettyNum(",")

total_licenses_all <- 
  STA %>% 
  st_drop_geometry() %>% 
  nrow()

bb_licenses <- round(sum(!STA$entire_apt, na.rm = TRUE) * 0.1)

total_licenses_STA <- total_licenses_all - bb_licenses
  
p_registration_compliance <- 
  (total_licenses_STA/ reg_property %>% nrow()) %>% scales::percent(accuracy = 0.1)

valid_STA <-
  STA %>% 
  st_drop_geometry() %>% 
  filter(!is.na(expiry)) %>% 
  filter(status== "Active" | status == "Expired") %>% 
  nrow()

n_valid_STA_bc <- 
  STA %>% 
  st_drop_geometry() %>% 
  filter(!is.na(expiry)) %>% 
  filter(status == "Active" | status == "Expired") %>% 
  filter(created <= end_reg) %>% 
  nrow()

# Number of licenses per ward 
valid_STA_bc <- 
  STA %>% 
  filter(!is.na(expiry)) %>% 
  filter(status== "Active" | status == "Expired") %>% 
  filter(created <= end_reg)


# Find property_IDs within the buffer of each license
ids <- 
  valid_STA_bc %>% 
  st_buffer(dist = 200) %>% 
  st_intersection(.,reg_property) %>% 
  st_drop_geometry() %>% 
  group_by(id) %>% 
  summarize(property_IDs = list(property_ID))

ids_sta_pid <- 
  ids %>% 
  nrow()

unique_PIDs <- 
  valid_STA_bc %>% 
  st_buffer(dist = 200) %>% 
  st_intersection(.,reg_property) %>% 
  st_drop_geometry() %>% 
  distinct(property_ID, .keep_all = TRUE) %>% 
  nrow()

# Number of license application by ward
picton_n_license <- 
  valid_STA_bc %>% 
  st_join(EW, .) %>% 
  st_drop_geometry() %>% 
  count(ward) %>% 
  arrange(desc(n)) %>% 
  slice(1) %>% 
  pull(n)

hallo_n_license <- 
  valid_STA_bc %>% 
  st_join(EW, .) %>% 
  st_drop_geometry() %>% 
  count(ward) %>% 
  arrange(desc(n)) %>% 
  slice(2) %>% 
  pull(n)

well_n_license <- 
  valid_STA_bc %>% 
  st_join(EW, .) %>% 
  st_drop_geometry() %>% 
  count(ward) %>% 
  arrange(desc(n)) %>% 
  slice(3) %>% 
  pull(n)

total_n_license <- 
  valid_STA_bc %>% 
  st_drop_geometry() %>% 
  nrow()


```

On 9 October 2018, Prince Edward County's city council approved a bylaw regulating STAs in its territory. The bylaw require STA hosts to acquire one license per accommodation, to be renewed annually, with license fees depending on the number of guestrooms within the accommodation. The County recognizes four different types of STAs: traditional B&Bs, where the owner lives on site; partial or entire principal-residence STAs; partial or entire secondary-residence STAs; and on-farm tourist homes. The bylaw does not set any restrictions on any of these categories, but it sets a maximum density of STAs for each zone or area of the County, and limits the total number of STAs County-wide to 15% of existing dwelling units. The licensing system came into effect in November 2019; as of that date, all STAs needed to be registered with the County. Covid-related travel restrictions were imposed in March 2020, and the County issued a moratorium on STA licensing on 30 September 2020. STA registration was reinstated for traditional B&Bs and principal-residence STAs on 22 June 2021, but the moratorium has remained in place for secondary-residence STAs, which therefore are currently unable to be licensed. The following chapter provides an overview of compliance with the STA licensing system, and the possible impact of the licensing moratorium on the reduced supply of STAs during the pandemic.


## STA license uptake and compliance

```{r daily_compliance}

dates <- as.Date(as.Date("2020-01-01"):max(daily$date), origin = "1970-01-01")

active_license <- map_dfr(dates, ~{
  total <- 
    STA |> 
    filter(status == "Active", created <= .x, (entire_apt | runif(n()) > 0.1))
  
  n_1 <- nrow(total)
  n_2 <- total |> 
    filter(expiry >= .x) |> 
    nrow()
  
  tibble(date = .x, category = c("Active, possibly expired",
                                 "Active, not expired"),
         n = c(n_1, n_2))
})

compliance <- 
  daily %>% 
  filter(status != "B", date >= "2020-01-01") %>% 
  count(date) |> 
  mutate(category = "STA", .before = n) |> 
  bind_rows(active_license) |> 
  arrange(date, category)

compliance_perc <- 
  compliance |> 
  group_by(date) |> 
  summarize(pos_exp_pct = n[category == "Active, possibly expired"] / n[category == "STA"],
            not_exp_pct = n[category == "Active, not expired"] / n[category == "STA"]) |> 
  summarize(across(c(pos_exp_pct, not_exp_pct), ~scales::percent(mean(.x), 0.1)))
  

```

```{r make_fig_3_1}

figure_3_1_fun <- function(regular = "", condensed = "") {

compliance %>% 
  group_by(category) %>%
  mutate(n = slide_dbl(n, mean, .before = 6, .complete = TRUE)) %>%
  ungroup() %>%
  mutate(label = if_else(date == "2021-07-01", category, NA_character_)) %>% 
  ggplot() +
  geom_line(aes(date, n, colour = category), size = 1) +
  geom_label(aes(date, n, label = label, colour = category)) +
  scale_color_manual(values = col_palette[1:3]) +
  theme_minimal() +
  ylab("") +
  xlab("") +
  theme(legend.title = element_blank(), legend.position = "none",
        panel.grid.minor.x = element_blank())
  
}

if (build_figures) {
  ggsave(here("output", "figures", "figure_3_1.pdf"), 
         plot = figure_3_1_fun(), 
         width = 9, height = 4.2, units = "in", useDingbats = FALSE)
  
  extrafont::embed_fonts(here("output", "figures", "figure_3_1.pdf"))
}

```

Since the start of the registration in November 2019, a total of `r n_reg_property` listings have been posted on Airbnb or Vrbo in Prince Edward County. During the same time period, there have been `r total_licenses_all` license applications. Approximately `r bb_licenses` of these applications were for traditional B&Bs, leaving `r total_licenses_STA` applications for STAs on Airbnb and other online platforms. In theory, each of listings should be registered, which suggests an overall compliance rate of `r p_registration_compliance`. In reality compliance is likely somewhat better than this, since there will be some duplicate listings on STA platforms which only correspond to a single STA registration (although we have attempted to detect and remove any such cases already), but broadly speaking it is possible to conclude that a weak majority of STA hosts have either complied or are in the process of complying with the registration system. A somewhat more positive perspective emerges from comparing the number of active STAs each day with the number of active licenses on that same day (Figure \@ref(fig:fig-3-1)). On average each day in 2020 and 2021, `r compliance_perc$not_exp_pct` of active listings were potentially operating with an active license, or `r compliance_perc$pos_exp_pct` if expired licenses are considered still valid. (License renewals were not halted along with new licenses as part of the registration moratorium, so existing STA hosts should have renewed their licenses, but given the general confusion of the pandemic, it would be understandable if some hosts who had previously been correctly engaging with the registration system misunderstood what was required of them during the moratorium period.)

``` {r fig-3-1, include = TRUE, fig.cap = '(ref:fig-3-1)', fig.align = "center"}

figure_3_1_fun()

```
(ref:fig-3-1) _Active STA licenses in circulation compared to active STRs_

Out of the `r total_licenses_all` total applications received by the County, `r valid_STA` were approved, with `r n_valid_STA_bc` of them occurring prior to the moratorium on STA licenses. The ward that received the most applications was Picton with `r picton_n_license`, followed by Hallowell with `r hallo_n_license`, and Wellington with `r well_n_license` (Table \@ref(tab:tab-3-1)). `r ids_sta_pid` of the `r total_n_license` short-term accommodations with valid licenses had one or more STR property within a 200-meter buffer of their address (Airbnb and Vrbo obfuscate listing locations by shifting them randomly up to 200 m on their websites), which strongly suggests that most licensed properties are in fact hosting an active STA. 

``` {r tab-3-1, include = TRUE}

library(kableExtra)

STA %>% 
  filter(!is.na(expiry)) %>% 
  filter(status== "Active" | status == "Expired") %>% 
  filter(created <= end_reg) %>%
  st_join(EW, .) %>% 
  count(ward) %>% 
  st_drop_geometry() %>% 
  add_row(ward = "Prince Edward County",
          n = n_valid_STA_bc) %>% 
  arrange(desc(n)) %>% 
  rename(Ward = ward,
         `STA licenses issued` = n) %>% 
  kbl(caption = paste0("Number of active STA licenses by ward, 4 Nov. 2019 to 4 Sept. 2021"),
      align = "lrrrr") %>%
  kable_styling(latex_options = "scale_down") %>% 
  row_spec(1, background = paste0(col_palette[2], "50"))
  
```

```{r STA_portrait}

# Entire apt vs not entire apt
n_entire_apt <- 
  valid_STA_bc %>% 
  st_drop_geometry() %>% 
  filter(!is.na(entire_apt)) %>% 
  count(entire_apt) %>% 
  filter(entire_apt == TRUE) %>% 
  pull(n)

n_not_entire_apt <- 
  valid_STA_bc %>% 
  st_drop_geometry() %>% 
  filter(!is.na(entire_apt)) %>% 
  count(entire_apt) %>% 
  filter(entire_apt == FALSE) %>% 
  pull(n)

```

Out of the `r n_valid_STA_bc` STAs operating with a license during the 10 months in which the registration system was functioning, `r n_entire_apt` licenses were for entire-home properties, or housing units entirely operated as a STA. Most licensed accommodations (`r n_not_entire_apt`) were not entire housing units, but rather a portion of a property or a traditional short-term accommodation, such as a hotel or a camp site. One third (`r scales::percent(mean(STA$bedrooms == 3), 0.1)`) of licenses were issued for accommodations with 3 bedrooms. Close to a quarter (`r scales::percent(mean(STA$bedrooms == 2), 0.1)`) were for 2-bedroom accommodations and more than a quarter (`r scales::percent(mean(STA$bedrooms >= 4), 0.1)`) were for 4-or-more-bedroom accommodations. 


## Was the STA registration moratorium responsible for the drop in STA supply during the pandemic?

The previous two chapters have identified a strange pattern in Prince Edward County's STA market since the start of the pandemic. On the one hand, nightly prices have skyrocketed in relation to pre-pandemic trends, which is a clear symptom of high demand for STAs. But on the other hand, the number of active STAs has declined in relation to pre-pandemic trends. In a relatively unconstrained market, rising prices would be expected to be followed by increasing supply, as more sellers see an economic opportunity to take advantage of, which would in turn reduce prices through competition until supply and demand reach a new equilibrium. The fact that this has not happened in the County's STA market strongly suggests that an external factor has limited new STA supply.

Given the previous discussion, there are two plausible candidates for this external factor. First, the fact that the County has put a moratorium on STA registrations means that it has been illegal to operate a new STA since September 2020. Potential STA hosts may have decided to delay or cancel their plans to open a new STA because they wish to follow the law. We can test this possibility by examining the creation of new STA listings before and after the moratorium came into effect; if the moratorium is a significant driver of STA host behaviour, there will be a disconinuity in this behaviour around the moratorium start date. Second, it is possible that possible new STAs have been crowded out by second-home sales. This could occur if potential STA hosts have been unable to purchase properties to convert to STAs because, even at the higher prevailing nightly prices, they cannot justify the prices which second-home purchasers are willing to pay. We can test this possibility by comparing STA nightly prices with home sales prices to see if STAs are becoming more or less lucrative, relative to the cost of housing.

```{r created}

created_dates <- 
  property %>% 
  st_drop_geometry() %>% 
  filter(created >= "2019-01-01") %>% 
  mutate(period = case_when(
    created < start_reg ~ "pre-reg",
    created < start_mora ~ "reg",
    created < end_mora ~ "moratorium",
    created >= end_mora ~ "post-moratorium"
  )) %>% 
  group_by(period) %>% 
  summarize(first_date = min(created),
            last_date = max(created), 
            tot_n = n(),
            avg = n() / as.numeric(max(created) - min(created) + 1))

created_table <- 
  property %>% 
  st_drop_geometry() %>% 
  count(created) %>% 
  filter(created >= start_2019)

created_table <- 
  tibble(
  created = as.Date(as.Date("2019-01-01"):as.Date("2021-09-30"), 
                    origin = "1970-01-01")) |> 
  left_join(created_table) |> 
  replace_na(list(n = 0)) |> 
  mutate(period = case_when(
    created < start_reg ~ "pre-reg",
    created < start_mora ~ "reg",
    created < end_mora ~ "moratorium",
    created >= end_mora ~ "post-moratorium"))

created_avg <- 
  created_table |> 
  group_by(period) |> 
  summarize(avg = scales::comma(mean(n), 0.01))

created_mora <- 
  created_avg |> 
  filter(period == "moratorium") |> 
  pull(avg)

created_reg <- 
  created_avg |> 
  filter(period == "reg") |> 
  pull(avg)

created_pre <- 
  created_avg |> 
  filter(period == "pre-reg") |> 
  pull(avg)

created_post <- 
  created_avg |> 
  filter(period == "post-moratorium") |> 
  pull(avg)

```

```{r make_fig_3_2}

figure_3_2_fun <- function(regular = "", condensed = "") {
  
created_table |> 
  mutate(n = slide_dbl(n, mean, .before = 13)) |> 
  ggplot() +
  annotate("rect", xmin = start_reg, xmax = end_reg, fill = col_palette[1],
           ymin = -Inf, ymax = Inf, alpha = 0.2) +
  annotate("rect", xmin = start_mora, xmax = end_mora, fill = col_palette[2],
           ymin = -Inf, ymax = Inf, alpha = 0.2) +
  geom_label(aes(x = mean(c(start_reg, start_mora)), y = 8,
                 label = "STA registration\nin effect", fill = "white"), 
             colour = col_palette[1], inherit.aes = FALSE, hjust = 0.5) +
  geom_label(aes(x = mean(c(start_mora, end_mora)), y = 8,
                 label = "STA registration\nmoratorium", fill = "white"), 
             colour = col_palette[2], inherit.aes = FALSE, hjust = 0.5) +
  geom_line(aes(created, n, colour = period), size = 0.4) +
  geom_segment(aes(x = first_date, xend = last_date, y = avg, yend = avg,
                   colour = period, group = period), data = created_dates,
               lwd = 1) +
  scale_x_date(name = NULL) +
  scale_y_continuous(name = NULL, limits = c(0, 10)) +
  scale_colour_manual(values = col_palette[c(2, 3, 3, 1)]) +
  scale_fill_manual(values = "white") +
  theme_minimal() +
  theme(legend.position = "none", panel.grid.minor.x = element_blank())

}

if (build_figures) {
  ggsave(here("output", "figures", "figure_3_2.pdf"), 
         plot = figure_3_2_fun(), 
         width = 9, height = 4.2, units = "in", useDingbats = FALSE)
  
  extrafont::embed_fonts(here("output", "figures", "figure_3_2.pdf"))
}

```

Figure \@ref(fig:fig-3-2) shows the number of new properties created on Airbnb or Vrbo per day, both during the STA registration period and the STA registration moratorium. From the beginning of 2019 until the registration system came online, a daily average of `r created_pre` STA listings were created on Airbnb and Vrbo. This number dropped significantly, to `r created_reg`, during the 10-month period during which STA registrations were required. But, importantly, the rate of new listings was almost identical—`r created_mora`—during the subsequent registration moratorium. If the moratorium were actually dissuading new STAs from being created, there would be some evidence of this fact in the rate of listing creation. Instead, there is nothing to distinguish the rate of listing creation before and during the registration moratorium. This strongly suggests that the moratorium has not been an important factor in the decline in STA listings during the pandemic.

``` {r fig-3-2, include = TRUE, fig.cap = '(ref:fig-3-2)', fig.align = "center"}

figure_3_2_fun()

```
(ref:fig-3-2) _Number of new STR properties created by day, 7-day rolling average_

```{r STA_STR_analysis, include = T}

STA_by_status <- 
  STA %>%
  st_buffer(200) %>% 
  filter(created >= "2020-09-05") %>% 
  st_join(property %>% filter(created >= "2020-09-05")) %>% 
  st_drop_geometry() %>% 
  group_by(id, status, entire_apt) %>% 
  summarize(strs = list(property_ID)) 

n_STA_post_mora <- 
  STA_by_status %>% 
  nrow()

p_entire_home <- 
  STA_by_status %>% 
  group_by(entire_apt) %>% 
  summarize(n = n(), perc = n/n_STA_post_mora) %>% 
  filter(entire_apt == TRUE) %>% 
  pull(perc) %>% 
  scales::percent(accuracy = 0.1)

PID_STA <- 
  STA_by_status %>% 
  filter(!is.na(strs)) %>% 
  nrow()

p_STA_PID_active <- 
  STA_by_status %>% 
  filter(!is.na(strs)) %>% 
  group_by(status) %>% 
  summarize(n = n(), perc = n / PID_STA) %>% 
  filter(status == "Active") %>% 
  pull(perc) %>% 
  scales::percent(accuracy = 0.1)

PIDs_post_mora <- 
  STA %>%
  st_buffer(200) %>% 
  filter(created >= "2020-09-05") %>% 
  st_join(property %>% filter(created >= "2020-09-05")) %>% 
  st_drop_geometry() %>% 
  filter(!is.na(property_ID)) %>% 
  distinct(property_ID, .keep_all = TRUE) %>% 
  pull(property_ID)

PIDS_post_mora_total <- 
  daily %>% 
  filter(property_ID %in% PIDs_post_mora) %>% 
  group_by(property_ID) %>% 
  summarize(n_total = n())

PIDS_post_mora_status <- 
  daily %>% 
  filter(property_ID %in% PIDs_post_mora) %>% 
  group_by(property_ID, status) %>% 
  summarize(n_status = n()) %>% 
  left_join(PIDS_post_mora_total, by = "property_ID") %>% 
  mutate(perc = n_status / n_total)

grouped_status_PIDS_post_mora <- 
  PIDS_post_mora_status %>% 
  group_by(status) %>% 
  summarize(perc = mean(perc, na.rm = TRUE))

p_R_post_mora <- 
  grouped_status_PIDS_post_mora %>% 
  filter(status == "R") %>% 
  pull(perc) %>% 
  scales::percent(accuracy = 0.1)

```

A further piece of evidence can be found in the `r n_STA_post_mora` license applications that were received since the moratorium began, of which one third (`r p_entire_home`) were for STAs operated out of an entire housing unit (the remainder were for owner-occupied STAs or traditional STA accommodations such as B&Bs.) In slightly more than half of the cases (`r PID_STA` applications out of the `r n_STA_post_mora` total), we managed to identify at least one newly created STA listing that could be located at the address on the license application. (Because of the way that Airbnb and Vrbo disguise the exact location of listings, it is not possible to be certain.) Despite the registration moratorium, most of these newly-created properties were active during the moratorium, and yet only `r p_STA_PID_active` of the license applications these properties have been approved. On average, these STA listings have been reserved `r p_R_post_mora` of the nights since their creation. In sum, the evidence clearly suggests that the registration moratorium has not meaningfully impeded new STA activity in Prince Edward County.

```{r STA_sales_ratio}

sta_nightly_prices <-
  daily |> 
  filter(status == "R", listing_type == "Entire home/apt") |> 
  group_by(date) |> 
  summarize(nightly_price = mean(price)) |> 
  mutate(nightly_price = slide_dbl(nightly_price, mean, .before = 6)) |> 
  filter(date >= "2017-06-01")

sale_price <- 
  home_sales |> 
  st_drop_geometry() |> 
  group_by(sale_date) |> 
  summarize(avg_price = mean(sale_price, na.rm = TRUE)) |> 
  mutate(avg_price = slide_dbl(avg_price, mean, .before = 6)) |> 
  filter(sale_date >= "2017-06-01") |> 
  mutate(down_pay = avg_price * 0.2,
         P = avg_price - down_pay,
         amort = 25,
         mortgage_rate = 0.0175) |> 
  mutate(r = mortgage_rate / 12,
         n = amort * 12) |> 
  mutate(monthly_payments = P * r * (((1 + r) ^ n) / (((1 + r) ^ n) - 1))) |> 
  mutate(maintenance = 700,
         total_payments = (monthly_payments + maintenance) * 12,
         affordable_income_2021 = total_payments / 0.3) |> 
  select(date = sale_date, monthly_payments)

nights_booked_needed <- 
  sta_nightly_prices %>% 
  left_join(sale_price) %>% 
  mutate(nights_needed = monthly_payments / nightly_price) 

needed_to_afford_2018 <- 
  nights_booked_needed %>% 
  filter(date >= "2018-01-01", date < "2019-01-01") %>%
  summarize(mean(nights_needed, na.rm = TRUE)) %>% 
  pull() %>% 
  round(., digits = 2)

needed_to_afford_2021 <- 
  nights_booked_needed %>% 
  filter(date >= "2021-01-01") %>%
  summarize(mean(nights_needed, na.rm = TRUE)) %>% 
  pull() %>% 
  round(., digits = 2)

```

```{r make_fig_3_3}

figure_3_3_fun <- function(regular = "", condensed = "") {
  
  nights_booked_needed %>%
    ggplot() + 
    geom_line(aes(date, y = nights_needed), size = 1, color = col_palette[1],
              alpha = 0.5) + 
    stat_smooth(aes(date, y = nights_needed), se = FALSE, method = "loess",
                span = 0.4, color = col_palette[1]) +
    ylab(NULL) +
    xlab(NULL) +
    theme_minimal() +
    theme(panel.grid.minor.x = element_blank(),
          text = element_text(face = "plain", family = regular))

}

if (build_figures) {
  ggsave(here("output", "figures", "figure_3_3.pdf"), 
         plot = figure_3_3_fun(), 
         width = 9, height = 4.2, units = "in", useDingbats = FALSE)
  
  extrafont::embed_fonts(here("output", "figures", "figure_3_3.pdf"))
}

```

If registration and the moratorium do not appear to have had a direct effect, what about the role of housing prices in the decline of STA activity? We compared monthly mortgage payments (calculated with the methodology explained in the previous chapter) with average per-night prices for STAs. In 2018, on average, an STA needed to have `r needed_to_afford_2018` nights booked per month to afford mortgage monthly payments, relative to the prices of houses sold in the same year. This number has since been steadily increasing, reaching `r needed_to_afford_2021` nights in 2021. STA nightly prices were rather stable before the pandemic, and Figure \@ref(fig:fig-3-3) shows that the profitability of STA was declining as housing prices were slightly rising. While after the pandemic, a property operating as an STA makes much more money, the increases in average prices per night still cannot keep up with the sales prices of second homes.

``` {r fig-3-3, include = TRUE, fig.cap = '(ref:fig-3-3)', fig.align = "center"}

figure_3_3_fun()

```
(ref:fig-3-3) _Average number of reserved STA nights needed to afford monthly mortgage payment_


TKTK evidence of a new surge in STA activity:
 (A potential complication is that, since the moratorium was partially lifted in June 2021, an average of `r created_post` new listings have been created each day. Given that dedicated, entire-home STAs remain subject to the moratorium, the most likely expl)


\newpage
