# Moving forward: short-term accommodation regulation pathways in Prince Edward County

```{=html}
<style type="text/css">
  body{
  font-family: Futura, Helvetica, Arial;
}
</style>
```
<br>

```{r setup, include = FALSE, echo = FALSE}

library(here)
source(here("R", "01_startup.R"))

knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(include = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)

# Function for checking file modifications
mtime <- function(files) lapply(Sys.glob(files), function(x) file.info(x)$mtime)

# Set this to TRUE to output PDF figures
build_figures <- TRUE

qload(here("output", "/geometry.qsm"))
qload(here("output", "/str_processed.qsm"))
full_time_listings <- qread("output/full_time_listings.qs")
STA <- qread("output/STA.qs")
rent_sept <- qread("output/rent_sept.qs")

options(tinytex.verbose = TRUE)

library(extrafont)
extrafont::loadfonts(quiet = TRUE)


```

**We evaluate scenarios for the reinstatement of Prince Edward County's short-term accommodation registration, in light of the housing situation and registration analysis, ranging from restrict the number of STA licenses emitted to allowing an unlimited amount of STA licenses for residents operating out of their primary residence. Two geographic scopes are suggested: one that would include only the towns, and one that would encompass all of Prince Edward County. The restrict scenarios would permit the issuance of between 200 or 800 STA licenses, depending on desired market size. The redistribute scenarios could benefit up to 1,700 home sharers operating for a few weeks during summer months. Financial incentives to entice STA operators to keep their listings onto the long-term rental market would not be a viable option for Prince Edward County as it would require up to \$2,520 in rent supplement, depending on location.**

## What are the key issues?

On the basis of the previous analysis, this chapter provides regulatory pathways for the reinstatement of a STA registration system in Prince Edward County. We identify a set of generic best practices which are likely to improve any regulatory system, and then lay out two reform models: a "Redistribute" model that entails redirecting STA activity away from dedicated commercial operations and towards home sharing instead, and a "Restrict" model that limits the number of STAs which are allowed to operate in order to limit the number of homes which are converted into STAs. Before we discuss our recommendations for regulatory reform of STAs in Prince Edward County, though, it is important to be clear about the key housing problems facing the County, and the extent to which regulating STAs can be part of the solution.

### Covid-19 is the most important driver of housing issues

Prince Edward County had been on a worsening housing affordability trajectory well since before the Covid-19 pandemic began, but the recent skyrocketing purchase prices and residential rents are a direct consequence of the pandemic. The preceding chapters suggest that the most important single factor in the pandemic-era deterioration of housing affordability is the purchase of second homes. This fact introduces a considerable amount of uncertainty into forward-looking regulatory reforms. The pandemic will not last forever; when it ends, will housing price growth slow down? Will it in fact reverse?

The additional complexity with regard to STAs is that, as we will discuss momentarily, the evidence is clear that second-home demand has come at the expense of STA supply. If demand for second homes declines after the pandemic, that might signal a decline in demand for STAs as well, or it might shift the financial conditions of investment properties to make STAs a more attractive investment, driving a new surge in STA activity going forward. There is no easy answer here; the point is simply to underscore the high level of uncertainty in discussing Prince Edward County's post-pandemic housing market.

### STAs and second homes are a single market

In major urban markets, research has established a high degree of substitutability between STAs and tenant-occupied apartments (Barron et al. 2020). Landlords face an easier decision transitioning between renting long-term to tenants and renting short-term to tourists than owner-occupiers do, since the latter would need to find a new home before they could convert their existing home into an STA. The consequence of this fact is that marginal changes in STA supply in urban areas tend to add or subtract from existing long-term rental housing. If the economic advantages of STAs increase, we would expect to see existing long-term rentals converted to STAs. And if regulations restrict STAs, we would expect to see former STAs converted to long-term rentals.

An implication of our analysis in this report is that, in Prince Edward County, the comparable substitutability is between STAs and second homes. The spike in demand for second homes since 2020 has reduced the supply of STAs relative to the pre-pandemic trend. And---importantly for future regulatory prospects---under current market conditions, the most likely outcome of restricting STA operations will be to increase second-home sales, as properties are converted from STA to second home.

### More (multifamily) housing is needed

The basic fact of the County's curent housing dilemma is that the pandemic has caused a surge in non-local housing demand which local supply can only meet through dramatic price increases. Even if local County residents vastly outnumber non-local second home purchasers, the latter's demand will drive prices at the margins far beyond the ability of local County incomes to afford. This will be true across nearly the entirety of the residential spectrum: seasonal service workers will be priced out of rental apartments, but even well-paid professionals such as doctors will be priced out of the ownership market. As long as high non-local demand remains in place, there will be no easy answers to housing affordability questions.

While in general the simplest answer to spiking demand for a commodity (in this case housing) is to add more supply, this is a long-term solution to a problem which has an important short-term component. However, to the extent that additional housing supply is permitted in the County, one promising avenue is to direct that supply to housing typologies which are better suited for local demand than for non-local demand. At the top of this list would be multifamily housing, and particularly multifamily purpose-built rental housing in Picton and Wellington. Rental apartments in these areas would help meet the need for housing for service workers while being less likely to be diverted into the second-home market.


## Regulatory best practices

Below we discuss specific regulatory scenarios that make different trade-offs in the STA market. However, there are a number of regulatory best practices which are likely to be worth implementing regardless of which regulatory scenario is followed.

### Prioritize owner-occupied STAs

First, regardless of how strict or permissive an STA regulatory system is, it should differentiate between STAs operated out of primary residences ("home sharing") and STAs operated out of secondary residences (commercial operations), and prioritize the former. Numerous studies that have showed that the trade-off between STAs and housing affordability is largest with dedicated STAs (Wachsmuth et al 2021; Garcia-López et al. 2019, Wachsmuth & Weisler 2018; Horn & Merante 2017). In city and town contexts, other detrimental effects of dedicated STAs include noise, littering and overcrowding (Celata and Romano 2020; Nieuwland and van Melik 2020). So, to the extent that the County contemplates restricting STA activities, restrictions should be targeted at commercial STAs and thereby allow STA activity in the County to be redirected to owner-occupied, homesharing STAs.

### Make display of license numbers mandatory

Effective STA registration systems require mandatory licensing of all listings. Any loopholes in licensing create grey areas for enforcement which, in practice, are difficult and time-intensive to solve. If a property is not licensed, does that mean that it is exempt from licensing, or that it is in violation? If enforcement officers have to answer that question every time a potential enforcement action is being contemplated, scarce enforcement resources will be stretched thin. Prince Edward County's STA regulations already require mandatory licensing, but they are missing an additional feature which allows the benefits of mandatory licensing to be most easily realized, which is mandatory display of license numbers in all online advertisements for STAs.

There are two reasons why this is a crucial feature of a registration system. First, it allows municipalities to connect the anonymous, online identities of STA hosts with the real-life identifies of property owners and property managers. If registration is mandatory but license numbers are not universally displayed, then the municipality faces the same incomplete information problem as when licensing is not fully mandatory.

Second, mandatory license display allows the regulation of STAs to shift away from the land use to the advertisement of a rental, which dramatically simplifies enforcement options. If, for example, STA licenses are business licenses which carry the condition of displaying the license number, then failure to display a license number is a violation which does not require an in-person inspection to substantiate. There will of course be other enforcement issues that will require visits from enforcement officers, but ensuring the basic integrity of the registration system will become much easier.

### Reduce grandfathering of existing licenses where possible

Subsection 34(9) of the Ontario Planning Act establishes a right to legal nonconforming use---or "grandfathering"---when the zoning bylaw is amended to prohibit a previous valid land use. To the extent that STAs are regulated through the zoning bylaw, grandfathering sets important limitations on the effectiveness of future policy changes. In particular, any change to the STA regulatory landscape which makes it more difficult to operate STAs will enormously benefit current STA operators, ultimately reducing the efficacy of the rules. Since the County's current STA regulations are enacted through the zoning bylaw, all existing licensed STAs will in theory be able to operate indefinitely, including after property sales, as legal nonconforming uses.

The County should take reasonable steps to moderate the impact of grandfathering on its ability to set appropriate STA policy going forward. One promising avenue is to shift STA regulation out of land use and into business licensing. If a business license is required to advertise or operate an STA, this could provide the County with more flexibility about which conditions it chooses to impose on the granting and renewal of these licenses in a manner which sidesteps grandfathered land uses. In particular, if a license is required to _advertise_ an STA online, this is not a land-use question, and legal nonconforming uses would not be a relevant concept.

## Two regulatory reform scenarios for Prince Edward County

Based on the findings presented earlier in the report, we identified two plausible models for reforming Prince Edward County's STA registration system, which we call "Redistribute" and "Restrict". The "Redistribute" model entails adding a principal-residence requirement to the STA license, and thus implies redirecting STA activity away from dedicated commercial operations and towards home sharing instead. The "Restrict" model entails leaving the existing relatively unconditional license in place, but limiting the number of licenses which are issued, in order to limit the number of homes which are converted into STAs.

Both of these models aim to preserve the important benefits of the tourist accommodation sector to the County's economy, while reducing the negative housing-affordability impacts of STAs. But they make different trade-offs to achieve these goals. The "Redistribute" model will be more disruptive to existing commercial STA operators, whose business model will no longer be permitted, but will consequently return more housing units to the long-term market and also provide much greater economic development opportunities for existing County residents. The "Restrict" model will be much less disruptive to existing STA operators, but as a result will have more modest impacts on housing supply.

For each of these two models, we modeled a scenario where the rules apply to the entire County ("towns + townships") and a scenario where the rules apply only in the towns, for a total of four regulatory possibilities.


## Model #1: "Redistribute"

The "Redistribute" model aims to redirect STA activity in Prince Edward County away from full-time, dedicated STAs which remove housing from the long-term market towards part-time, home-sharing STAs which allow local residents to share in the economic benefits of tourism accommodation. The model achieves this goal by adding a principal-residence requirement to the STA license: only hosts operating out of their own home will be allowed to operate STAs.

There are two important implications of the "Redistribute" model. The first is that there will be _more_ STA hosts than before, since a smaller number of full-time STAs will be replaced with a larger number of part-time STAs. However, because these hosts will be operating STAs out of their own homes, STAs will no longer be in a zero-sum relationship with housing supply, and thus the negative housing-market impacts of STAs will be significantly reduced.

The second implication of the "Redistribute" model is that the positive economic benefits of STAs will be much more equitably distributed among County residents. Currently, a substantial share of the County's STA listings are owned by non-locals. (It is impossible to measure this precisely, but as an absolutely minimal estimate, around 1 in 8 STA listings in the County is operated by a host which also operates listings outside the County.) This means that a substantial share of STA revenue leaves the County. Moreover, even for STAs which are locally owned, STA revenue is highly concentrated among a small number of operators—10% of STA hosts earn approximately half of all the revenue. If STAs can only be operated out of principal residences, this means that there will be no non-local STAs at all—so all STA revenue will stay in the County—and that revenue will be distributed more evenly among a larger number of part-time hosts. Finally, a shift to proportionately more home-sharing STAs would provide more opportunities for local firms to provide intermediary services such as property management and cleaning to a larger pool of STA hosts.

### "Redistribute" (towns + townships)

```{r redistribute_town_townships, include = FALSE}

pool_tt <- 
  daily %>% 
  filter(date >= summer_start_2021, status == "R") %>% 
  filter(property_ID %in% (full_time_listings %>% 
                             filter(year_season == "Summer 2021") %>% 
                             pull(property_ID)))

total_nights_tt <- 
  pool_tt %>% 
  nrow() %>% 
  round(digit = -3)

revenue_summer_2021_tt <- 
  pool_tt %>% 
  summarize(sum(price)) %>% 
  pull() %>% 
  scales::dollar(0.1, 1 / 1000000)

new_home_sharers_tt <- 
  (nrow(pool_tt) / 28) %>% 
  round(digit = -2)

earnings_per_cap_tt <- 
  {(pool_tt %>% summarize(sum(price)) %>% pull())/new_home_sharers_tt} %>% 
  round(digit = -2) %>% 
  prettyNum(",") %>% 
  paste0("$", .)

grandfathered_tt <- 
  STA %>% 
  filter(status == "Active", entire_apt) %>% 
  nrow()

avg_nb_nights_freh_tt <- 
  pool_tt %>% 
  group_by(property_ID) %>% 
  summarize(nb_nights = n()) %>% 
  summarize(mean(nb_nights)) %>% 
  pull()

total_nights_after_gf_tt <- 
  {total_nights_tt - (grandfathered_tt * avg_nb_nights_freh_tt)} %>% 
  round(digit = -3)

revenue_summer_2021_after_gf_tt <- 
  {(pool_tt %>% 
      summarize(sum(price))) / nrow(pool_tt) * 
      (total_nights_tt - (grandfathered_tt * avg_nb_nights_freh_tt))} %>% 
  pull() %>% 
  scales::dollar(0.1, 1 / 1000000)

new_home_sharers_after_gf_tt <- 
  ((total_nights_tt - (grandfathered_tt * avg_nb_nights_freh_tt)) / 28) %>% 
  round(digit = -2)

earnings_per_cap_after_gf_tt <- 
  ({(pool_tt %>% 
       summarize(sum(price))) / total_nights_tt * 
      (total_nights_tt - (grandfathered_tt * avg_nb_nights_freh_tt))} /
     ((total_nights_tt - (grandfathered_tt * avg_nb_nights_freh_tt)) / 28)) %>% 
  pull() %>%
  scales::dollar(100)
  
```
```{r make_fig_4_1}

## Fonts don't work properly, so this code needs to be run from RStudio to generate figure

# remotes::install_github("hrbrmstr/waffle", force = TRUE)
# library(waffle)
# library(hrbrthemes)
# library(patchwork)
# 
# full_time_tt <- 
#   full_time_listings %>% 
#   filter(year_season == "Summer 2021") %>% 
#   nrow()
# 
# home_sharing_tt <- 
#   daily %>% 
#   filter(date >= summer_start_2021, status == "R") %>% 
#   filter(!property_ID %in% (full_time_listings %>% 
#                               filter(year_season == "Summer 2021") %>% 
#                               pull(property_ID))) %>% 
#   distinct(property_ID) %>% 
#   nrow()
# 
# waffle_df <-
#   tibble(scenario = "Current market", group = "Current full-time", 
#          value = full_time_tt) |> 
#   add_row(scenario = "Current market", group = "Current home sharing", 
#           value = home_sharing_tt) |> 
#   add_row(scenario = "Redistribute (towns + townships)", 
#           group = "New home sharing", value = new_home_sharers_tt) |> 
#   add_row(scenario = "Redistribute (towns + townships)", 
#           group = "Current home sharing", value = home_sharing_tt) |> 
#   add_row(scenario = "With grandfathering", group = "Current full-time",
#           value = grandfathered_tt) |> 
#   add_row(scenario = "With grandfathering", group = "New home sharing",
#           value = new_home_sharers_after_gf_tt) |> 
#   add_row(scenario = "With grandfathering", group = "Current home sharing",
#           value = home_sharing_tt) |> 
#   mutate(value = value / 10) |> 
#   mutate(group = factor(
#     group, levels = c("Current full-time", "New home sharing", 
#                       "Current home sharing")))
# 
# cols <- c("Current full-time" = col_palette[6],
#           "New home sharing" = col_palette[2], 
#           "Current home sharing" = col_palette[5])
# 
# labels <- c("Current full-time" = "home",
#             "New home sharing" = "home", 
#             "Current home sharing" = "home")
# 
# waffle_df |> 
#   ggplot(aes(label = group, colour = group, fill = group, values = value)) +
#   geom_pictogram(size = 3.2, n_rows = 8) +
#   scale_label_pictogram(name = NULL, values = labels) +
#   scale_x_discrete(expand = c(0, 0)) +
#   scale_y_discrete(expand = c(0, 0)) +
#   scale_colour_manual(values = cols, guide = "none") +
#   scale_fill_manual(values = cols, guide = "none") +
#   guides(label = guide_legend(override.aes = list(colour = "green",
#                                                   fill = "blue"))) +
#   coord_equal() +
#   facet_wrap(vars(scenario), nrow = 3) +
#   theme_minimal() +
#   theme_enhance_waffle() +
#   theme(legend.position = "bottom", panel.background = element_blank(),
#         strip.text = element_text(size = 12, hjust = 0))
 
```

The most straightforward way to implement the "Redistribute" model would be to establish a principal-residence requirement for all STA licenses in the County. This would create a cascade of changes in the STA market which are impossible to predict in advance, but we can estimate the magnitude of the transformation by examining the potential impacts on a static approximation of the STA market in Summer 2021.

In Summer 2021, there were approximately `r prettyNum(total_nights_tt, ",")` reserved nights (earning `r revenue_summer_2021_tt` million) which occurred in commercially-operated STAs. If this activity were redistributed to principal-residence STAs, it could support `r scales::comma(new_home_sharers_tt)` home sharers operating between three to five weeks in the summer, each earning `r earnings_per_cap_tt` in additional annual income.

Some of the existing full-time STA hosts are operating with a valid license, and they might therefore be permitted to continue operating as a legal nonconforming use. Under the conservative assumption that each of the `r grandfathered_tt` currently active secondary-home licenses belongs to an active full-time STA, and thus these STAs would stay in operation, there would have been `r prettyNum(total_nights_after_gf_tt, ",")` reserved nights (earning `r revenue_summer_2021_after_gf_tt` million) which occurred in commercially operated STAs which would no longer be permitted. This activity could support `r scales::comma(new_home_sharers_after_gf_tt)` home sharers each earning `r earnings_per_cap_after_gf_tt` annually. Figure \@ref(fig:fig-4-1) summarizes potential changes in STAs under these scenarios.

(ref:fig-4-1) *Expected impact of redistribution regulations (in all the County) on the STA market*


### "Redistribute" (towns only)

```{r redistribute_town, include = F}
pool_t <- 
daily %>% 
    filter(ward %in% (EW %>% filter(type != "Township") %>% pull(ward))) %>% 
  filter(date >= summer_start_2021, status == "R") %>% 
  filter(property_ID %in% (full_time_listings %>% 
                             filter(year_season == "Summer 2021") %>% 
                             pull(property_ID)))

total_nights_t <- 
pool_t %>% 
  nrow() %>% 
  round(digit = -3)

revenue_summer_2021_t <- 
  pool_t %>% 
  summarize(sum(price)) %>% 
  pull() %>% 
  scales::dollar(0.1, 1 / 1000000)

new_home_sharers_t <- 
(nrow(pool_t) / 28) %>% 
  round(digit = -2) 

earnings_per_cap_t <- 
  {(pool_t %>% summarize(sum(price)) %>% pull()) / (nrow(pool_t) / 28)} %>% 
  round(digit = -2) %>% 
  prettyNum(",") %>% 
  paste0("$", .)

grandfathered_t <- 
  STA %>% 
    st_intersection((EW %>% filter(type != "Township"))) %>% 
  filter(status == "Active", entire_apt) %>% 
  nrow()

avg_nb_nights_freh_t <- 
pool_t %>% 
  group_by(property_ID) %>% 
  summarize(nb_nights = n()) %>% 
  summarize(mean(nb_nights)) %>% pull()

total_nights_after_gf_t <- 
  {nrow(pool_t) - (grandfathered_t*avg_nb_nights_freh_t)} %>% 
  round(digit = -3)

revenue_summer_2021_after_gf_t <- 
  {(pool_t %>% 
  summarize(sum(price))) / nrow(pool_t) * 
      {nrow(pool_t) - (grandfathered_t*avg_nb_nights_freh_t)}} %>% 
  pull() %>% 
  scales::dollar(0.1, 1 / 1000000)

new_home_sharers_after_gf_t <- 
({nrow(pool_t) - (grandfathered_t*avg_nb_nights_freh_t)} / 28) %>% 
  round(digit = -2)

earnings_per_cap_after_gf_t <- 
  (pull({(pool_t %>% 
  summarize(sum(price))) / nrow(pool_t) * 
      {nrow(pool_t) - (grandfathered_t*avg_nb_nights_freh_t)}} / 
    new_home_sharers_after_gf_t)) %>% 
  scales::dollar(100)
  
```

Because the negative housing-market impacts of dedicated STAs are particularly high in denser contexts where both permanent and seasonal workers are more likely to be looking for housing, one possibility is that the "Redistribute" model could be limited to Picton, Wellington and Bloomfield (what we call the "towns only" option). Under this scenario, in Summer 2021 there were `r prettyNum(total_nights_t, ",")` reserved nights (`r revenue_summer_2021_t` million in revenue) in commercial STAs. These could support `r new_home_sharers_t` new home sharers active for three to five weeks in the summer, for an average revenue of `r earnings_per_cap_t` per household. If the `r grandfathered_t` currently active second-home licenses in the towns were grandfathered in, the number of reserved nights being redistributed to households would be `r prettyNum(total_nights_after_gf_t, ",")` (`r revenue_summer_2021_after_gf_t` million in revenue) to support `r new_home_sharers_after_gf_t` home sharers, each earning `r earnings_per_cap_after_gf_t` annually. Figure \@ref(fig:fig-4-2) summarizes potential changes in STAs under these scenarios.

```{r make_fig_4_2}

### Fonts don't work

# full_time_t <-
#   property %>%
#   filter(property_ID %in% filter(full_time_listings, 
#                                  year_season == "Summer 2021")$property_ID) %>%
#   filter(ward %in% (EW %>% filter(type != "Township"))$ward) %>%
#   nrow()
# 
# home_sharing_t <-
#   daily %>%
#   filter(ward %in% (EW %>% filter(type != "Township") %>% pull(ward))) %>%
#   filter(date >= summer_start_2021, status == "R") %>%
#   filter(!property_ID %in% (full_time_listings %>%
#                               filter(year_season == "Summer 2021") %>%
#                               pull(property_ID))) %>%
#   distinct(property_ID) %>%
#   nrow()
# 
# waffle_df_t <-
#   tibble(scenario = "Current market", group = "Current full-time", 
#          value = full_time_tt) |> 
#   add_row(scenario = "Current market", group = "Current home sharing", 
#           value = home_sharing_tt) |> 
#   add_row(scenario = "Redistribute (towns only)",
#           group = "Current full-time", value = full_time_tt - full_time_t) |> 
#   add_row(scenario = "Redistribute (towns only)", 
#           group = "New home sharing", value = new_home_sharers_t) |> 
#   add_row(scenario = "Redistribute (towns only)", 
#           group = "Current home sharing", value = home_sharing_tt) |> 
#   add_row(scenario = "With grandfathering", group = "Current full-time",
#           value = full_time_tt + grandfathered_t - full_time_t) |> 
#   add_row(scenario = "With grandfathering", group = "New home sharing",
#           value = new_home_sharers_after_gf_t) |> 
#   add_row(scenario = "With grandfathering", group = "Current home sharing",
#           value = home_sharing_tt) |> 
#   mutate(value = value / 10) |> 
#   mutate(group = factor(
#     group, levels = c("Current full-time", "New home sharing", 
#                       "Current home sharing")))
# 
# cols <- c("Current full-time" = col_palette[6],
#           "New home sharing" = col_palette[2], 
#           "Current home sharing" = col_palette[5])
# 
# labels <- c("Current full-time" = "home",
#             "New home sharing" = "home", 
#             "Current home sharing" = "home")
# 
# waffle_df_t |>
#   ggplot(aes(label = group, colour = group, fill = group, values = value)) +
#   geom_pictogram(size = 3.2, n_rows = 8) +
#   scale_label_pictogram(name = NULL, values = labels) +
#   scale_x_discrete(expand = c(0, 0)) +
#   scale_y_discrete(expand = c(0, 0)) +
#   scale_colour_manual(values = cols, guide = "none") +
#   scale_fill_manual(values = cols, guide = "none") +
#   guides(label = guide_legend(override.aes = list(colour = "green",
#                                                   fill = "blue"))) +
#   coord_equal() +
#   facet_wrap(vars(scenario), nrow = 3) +
#   theme_minimal() +
#   theme_enhance_waffle() +
#   theme(legend.position = "bottom", panel.background = element_blank(),
#         strip.text = element_text(size = 12, hjust = 0))

```

(ref:fig-4-2) *Expected impact of redistribution regulations (only in Towns) on the STA market*


## Model #2: "Restrict"

The "Restrict" model aims to leave the existing structure of the STA market in Prince Edward County largely intact, but to limit the total number of STAs and thus also limit the number of homes which are converted into STAs. The model achieves this goal by leaving the currently permissive license criteria in place, but also setting a cap on the number of licenses which will be issued. Compared to the "Redistribute" model, this would be a smaller change to the status quo and a simpler program to administer, but it would also have a correspondingly smaller effect on the negative housing-market impacts of STAs, and would not be expected to produce the same local-economic-development benefits as "Redistribute". 

The "Restrict" model implies an STA market where there are fewer STA listings but these listings operate more intensively. This suggests that STA prices would continue on the high trajectory of the last 18 months, since external demand for STA accommodations will be channeled into a smaller number of offerings. It also implies that STA activity would be concentrated in a smaller number of hands, since the economic value of an STA license would be quite high. We would expect to see very few casual home sharers operating under the "Restrict" model, and would expect to see consolidations of existing STA operators into larger entities to realize greater economies of scale.

Under the "Restrict" model, it would be important to ensure that all STA licenses are being actively used, because defunct licenses would limit the municipality's ability to affect STA market outcomes by changing the license quotas. This would suggest a high-fee business model, where, for example, licenses require relatively extensive in-person inspections and supporting documentation, and cost recovery is correspondingly high.

```{r restrict, include = FALSE}

levels_2019_tt <- 
daily %>% 
  filter(listing_type == "Entire home/apt", 
         date >= summer_start_2021 - years(2), 
         date <= summer_end_2021 - years(2),
         status == "R") %>% 
  count(date, sort = TRUE) %>% 
  slice(1) %>%
  pull(n) %>% 
  round(digits = -1)

levels_2021_tt <- 
  daily %>% 
  filter(listing_type == "Entire home/apt", date >= summer_start_2021, 
         date <= summer_end_2021, status == "R") %>% 
  count(date, sort = TRUE) %>% 
  slice(1) %>%
  pull(n) %>% 
  round(digits = -1)

levels_2019_t <- 
  daily %>% 
  filter(ward %in% (EW %>% filter(type != "Township") %>% pull(ward))) %>% 
  filter(listing_type == "Entire home/apt", 
         date >= summer_start_2021 - years(2), 
         date <= summer_end_2021 - years(2),
         status == "R") %>% 
  count(date, sort = TRUE) %>% 
  slice(1) %>%
  pull(n) %>% 
  round(digits = -1)

levels_2021_t <- 
  daily %>% 
  filter(ward %in% (EW %>% filter(type != "Township") %>% pull(ward))) %>% 
  filter(listing_type == "Entire home/apt", date >= summer_start_2021, 
         date <= summer_end_2021, status == "R") %>% 
  count(date, sort = TRUE) %>% 
  slice(1) %>%
  pull(n) %>% 
  round(digits = -1)


```

The precise impacts of the "Restrict" scenario would depend on the number of licenses which the County issued. For example, setting a total cap of `r levels_2019_tt` licenses across the entire municipality would allow the STA market to function at its 2019 level with little to no improvement of housing-market outcomes. `r levels_2021_tt` licenses would future STA activity at  2021 levels. A smaller license quota would imply a shrinking STA market. If license amounts were only restricted in the towns, then `r levels_2019_t` licenses would allow 2019 levels of activity, and `r levels_2021_t` licenses would allow 2021 levels.

The current number of active secondary-residence STA licenses is well below any plausible forward-looking "Restrict" license quota, so the grandfathering of existing licenses would not be an impediment to implenting the "Restrict" model.


## Comparing the scenarios

The "Redistribute" and "Restrict" models represent two different ways of balancing the positive and negative aspects of STA activity in Prince Edward County. The former attempts to steer STA activity towards more casual home sharing, to reduce the number of housing units converted to dedicated STAs while offering a new economic opportunity for local residents. But it would be a relatively disruptive change to the existing STA market. The latter, by contrast, is a more modest intervention which largely preserves the existing STA market but reduces its total size in order to reduce its negative impacts. It is a simpler model, and it is also likely to have more modest effects.

Our judgment is that the "Redistribute" model represents the best balance of developing the STA markets' positive aspects and mitigating its negative aspects. And, while we have presented estimations for the impacts of both a "towns + townships" and a "towns only" version of these models, in practice the likely effect of a "towns only" regulatory model will be to displace most STAs to the townships, rather than affecting the overall size or character of the market.


## Less promising regulatory options

In addition to the two main "Redistribute" and "Restrict" regulatory options, we considered several others which have been proposed or implemented in other jurisdictions but which, in our opinion, are not plausible paths forward for Prince Edward County. We briefly discuss them here.

### Incentives to STA hosts?

```{r short_term_incentive}

num_bedrooms <- 
  property %>% 
  st_drop_geometry() %>% 
  mutate(bedrooms = if_else(bedrooms >= 3, 3, bedrooms)) %>% 
  select(property_ID, bedrooms)

FREH_properties <- 
  full_time_listings %>% 
  distinct(property_ID) %>% pull()

LTM_revenue <- 
  daily %>% 
  left_join(num_bedrooms) %>% 
  filter(date >= max(daily$date) - years(1), status == "R", listing_type == "Entire home/apt",
         property_ID %in% FREH_properties) %>%
  mutate(ward = if_else(!ward %in% c("Picton", "Wellington", "Bloomfield"), 
                        "Township", ward)) %>% 
  group_by(property_ID, ward, bedrooms) %>% 
  summarize(rev_LTM = sum(price)) %>% 
  ungroup()
  
avg_rent <- 
rent_sept %>% 
  group_by(ward) %>% 
  summarize(rent = mean(rent))

rev_to_weight_mean <- 
  LTM_revenue %>% 
  group_by(bedrooms) %>% 
  summarize(monthly_rev_LTM = mean(rev_LTM) / 12,
            weight = n())

rent_to_weight_mean <- 
  rent_sept %>% 
  group_by(bedrooms) %>% 
  summarize(monthly_rent = mean(rent),
            weight = n())

avg_monthly_rev_LTM <- 
weighted.mean(rev_to_weight_mean$monthly_rev_LTM,
              rent_to_weight_mean$weight) %>% 
  round(-1) %>% 
  scales::dollar()

avg_monthly_rent <- 
weighted.mean(rent_to_weight_mean$monthly_rent,
              rent_to_weight_mean$weight) %>% 
  round(-1) %>% 
  scales::dollar()

var_rev_rent <- 
{(weighted.mean(rev_to_weight_mean$monthly_rev_LTM,
              rent_to_weight_mean$weight) - 
weighted.mean(rent_to_weight_mean$monthly_rent,
              rent_to_weight_mean$weight)) /
weighted.mean(rent_to_weight_mean$monthly_rent,
              rent_to_weight_mean$weight)} %>% 
  scales::percent()


```

```{r on_water, cache = TRUE, cache.lazy = FALSE}

qload(here("output", "geometry.qsm"), nthreads = availableCores())

ocean <-
  read_sf(here("data", "shapefiles", "lhy_000h16a_e.shp")) %>%
  st_transform(32618)

ocean <-
  ocean %>%
  st_filter(select(city, -everything()))

river <-
  read_sf(here("data", "shapefiles", "lhy_000c16a_e.shp")) %>%
  st_transform(32618)

river <-
  river %>%
  st_filter(select(city, -everything()))

water <- rbind(select(river, -everything()), select(ocean, -everything()))

rm(ocean, river, city)
```

```{r make_fig_4_3}

weight_rent_sept <- 
rent_sept %>% 
  group_by(ward, bedrooms)  %>% 
  summarize(mean_rent = mean(rent),
            weight = n())

weight_revenue_ward <- 
LTM_revenue %>% 
  group_by(ward, bedrooms) %>% 
  summarize(mean_rev = mean(rev_LTM) / 12,
            weight = n())

# Weighted average to compare with rents, weight of STRs
LTM_revenue_ward <- 
tibble(ward = c("Bloomfield", "Wellington", "Township", "Picton"), mean_rev = c(weighted.mean(weight_revenue_ward[weight_revenue_ward$ward == "Bloomfield",]$mean_rev,
weight_revenue_ward[weight_revenue_ward$ward == "Bloomfield",]$weight),

weighted.mean(weight_revenue_ward[weight_revenue_ward$ward == "Wellington",]$mean_rev,
weight_revenue_ward[weight_revenue_ward$ward == "Wellington",]$weight),

weighted.mean(weight_revenue_ward[weight_revenue_ward$ward == "Township",]$mean_rev,
weight_revenue_ward[weight_revenue_ward$ward == "Township",]$weight),

weighted.mean(weight_revenue_ward[weight_revenue_ward$ward == "Picton",]$mean_rev,
weight_revenue_ward[weight_revenue_ward$ward == "Picton",]$weight)))

# Weight average to compare with revenue, weight of STRs
set_rent_ward <- 
tibble(ward = c("Bloomfield", "Wellington", "Township", "Picton"), mean_rent = c(weighted.mean(weight_rent_sept[weight_rent_sept$ward == "Bloomfield",]$mean_rent,
weight_revenue_ward[weight_revenue_ward$ward == "Bloomfield",]$weight),

weighted.mean(weight_rent_sept[weight_rent_sept$ward == "Wellington",]$mean_rent,
weight_revenue_ward[weight_revenue_ward$ward == "Wellington",]$weight),

weighted.mean(weight_rent_sept[weight_rent_sept$ward == "Township",]$mean_rent,
weight_revenue_ward[weight_revenue_ward$ward == "Township",]$weight),

weighted.mean(weight_rent_sept[weight_rent_sept$ward == "Picton",]$mean_rent,
weight_revenue_ward[weight_revenue_ward$ward == "Picton",]$weight)))

rent_wards <- 
EW %>% 
  mutate(ward = if_else(type == "Township", "Township", ward)) %>%
  group_by(ward) %>% 
  summarize() %>% 
  left_join(set_rent_ward)

surplus_rent <- 
LTM_revenue_ward %>% 
  left_join(rent_wards) %>% 
  mutate(surplus = round(mean_rev - mean_rent, -1)) %>% 
  st_as_sf()

figure_4_3_fun <- function(regular = "", condensed = "") {
  
townships <- 
surplus_rent[surplus_rent$ward == "Township",] %>% 
  ggplot() +
  geom_sf(data = province, colour = "transparent", fill = "grey93") +
  geom_sf(
    aes(fill = surplus),
    lwd = 0,
    colour = "white"
  ) +
  geom_sf(data = streets, colour = "white", alpha = 0.2) +
  geom_sf(data = water, fill = "white", colour = "transparent") +
  scale_fill_stepsn(
    name = "Monthly dif. btw avg. rent\nand avg. full-time STA revenue",
    colors = col_palette[c(5, 3, 4, 6)], 
                       na.value = "grey80",
                       limits = c(min(surplus_rent$surplus), 
                                  max(surplus_rent$surplus)),
                       oob = scales::squish, 
                       labels = scales::dollar)+
  gg_bbox(surplus_rent)+
  theme_void() +
  theme(
    legend.position = "bottom",
    legend.key.width= unit(1.5, 'cm')
  )+
  geom_sf_label(aes(label = scales::dollar(round(surplus))),
                nudge_y = 5000, nudge_x = -5000)+
  ggtitle("Townships")

picton <- 
  surplus_rent[surplus_rent$ward == "Picton",] %>% 
  ggplot() +
  # geom_sf(data = province, colour = "transparent", fill = "grey93") +
  geom_sf(
    aes(fill = surplus),
    lwd = 0,
    colour = "white"
  ) +
  geom_sf(data = streets, colour = "white", alpha = 0.2) +
  geom_sf(data = water, fill = "white", colour = "transparent") +
  scale_fill_stepsn(
    name = "Monthly dif. btw avg. rent\nand avg. full-time STA revenue",
    colors = col_palette[c(5, 3, 4, 6)], 
                       na.value = "grey80",
                       limits = c(min(surplus_rent$surplus), 
                                  max(surplus_rent$surplus)),
                       oob = scales::squish, 
                       labels = scales::dollar)+
  gg_bbox(surplus_rent[surplus_rent$ward == "Picton",])+
  theme_void() +
  theme(
    legend.position = "bottom",
    legend.key.width= unit(1.5, 'cm')
  )+
  geom_sf_label(aes(label = scales::dollar(round(surplus))))+
  ggtitle("Picton")

bloomfield <- 
  surplus_rent[surplus_rent$ward == "Bloomfield",] %>% 
  ggplot() +
  # geom_sf(data = province, colour = "transparent", fill = "grey93") +
  geom_sf(
    aes(fill = surplus),
    lwd = 0,
    colour = "white"
  ) +
  geom_sf(data = streets, colour = "white", alpha = 0.2) +
  geom_sf(data = water, fill = "white", colour = "transparent") +
  scale_fill_stepsn(
    name = "Monthly dif. btw avg. rent\nand avg. full-time STA revenue",
    colors = col_palette[c(5, 3, 4, 6)], 
                       na.value = "grey80",
                       limits = c(min(surplus_rent$surplus), 
                                  max(surplus_rent$surplus)),
                       oob = scales::squish, 
                       labels = scales::dollar)+
  gg_bbox(surplus_rent[surplus_rent$ward == "Bloomfield",])+
  theme_void() +
  theme(
    legend.position = "bottom",
    legend.key.width= unit(1.5, 'cm')
  )+
  geom_sf_label(aes(label = scales::dollar(round(surplus))))+
  ggtitle("Bloomfield")

wellington <- 
  surplus_rent[surplus_rent$ward == "Wellington",] %>% 
  ggplot() +
  # geom_sf(data = province, colour = "transparent", fill = "grey93") +
  geom_sf(
    aes(fill = surplus),
    lwd = 0,
    colour = "white"
  ) +
  geom_sf(data = streets, colour = "white", alpha = 0.2) +
  geom_sf(data = water, fill = "white", colour = "transparent") +
  scale_fill_stepsn(
    name = "Monthly dif. btw avg. rent\nand avg. full-time STA revenue",
    colors = col_palette[c(5, 3, 4, 6)], 
                       na.value = "grey80",
                       limits = c(min(surplus_rent$surplus), 
                                  max(surplus_rent$surplus)),
                       oob = scales::squish, 
                       labels = scales::dollar)+
  gg_bbox(surplus_rent[surplus_rent$ward == "Wellington",])+
  theme_void() +
  theme(
    legend.position = "bottom",
    legend.key.width= unit(1.5, 'cm')
  )+
  geom_sf_label(aes(label = scales::dollar(round(surplus))))+
  ggtitle("Wellington")

combined <- 
(townships + picton) / (bloomfield + wellington) & 
  theme(legend.position = "bottom")

combined + plot_layout(guides = "collect")

}

if (build_figures) {
  ggsave(here("output", "figures", "figure_4_3.pdf"), 
         plot = figure_4_3_fun(), 
         width = 8, height = 5, units = "in", useDingbats = FALSE)
  
  extrafont::embed_fonts(here("output", "figures", "figure_4_3.pdf"))
}

```

Another STA regulatory option that has been considered by municipalities in Canada (such as Toronto; Wachsmuth et al. 2021) and worldwide (deJong 2021) is a financial incentive to dedicated STA operators to entice them to rent their units on the long-term rental market. Under normal market conditions, STAs operating in strong markets are able to earn considerably more money than they could earn as long-term rentals. But, since the start of the Covid-19 pandemic, many of these markets have seen the gap between STA and long-term rental revenue narrow or even reverse, raising the question of whether it would be financially feasible to encourage STA hosts to return their units to the housing market.

To analyze the viability of financial incentives to STA hosts in Prince Edward County, we compare average STA revenues with average residential rents. To a first approximation, an effective financial incentive would need to roughly match the difference between monthly STA revenue and the long-term rents it would be possible to achieve in that housing unit. We approximate this difference by taking the average monthly revenue of full-time STAs over the last twelve months and average residential rents from September 2021, and weighing each proportionally to the number of bedrooms found in both STAs and rental units in the County. Across the County, full-time STA listings earned an average of `r avg_monthly_rev_LTM` a month, while the average monthly rent was of `r avg_monthly_rent` in September. This is a difference of `r var_rev_rent`.

```{r short_term_incentive_2}

subisdy_outside_wellington <- 
surplus_rent %>% 
  st_drop_geometry() %>% 
  filter(ward != "Wellington") %>% 
  summarize(mean(surplus)) %>% 
  pull() %>% 
  scales::dollar(10)

subisdy_wellington <- 
surplus_rent %>% 
  st_drop_geometry() %>% 
  filter(ward == "Wellington") %>% 
  summarize(mean(surplus)) %>% 
  pull() %>% 
  scales::dollar(10)

```

Figure \@ref(fig:fig-4-3) shows the average amount of monthly subsidy that would be necessary to induce an STA back onto the long-term market for each electoral ward. While average rents are similar across the County, monthly STA revenue differs substantially from one location to another. The only ward with substantially higher rents than average is Wellington. An incentive program would cost the County `r subisdy_wellington` per unit in Wellington each month, and `r subisdy_outside_wellington` in the rest of the county. The scale of this financial commitment makes it highly unlikely that an incentive program would be an effective use of public funds. The underlying cause is relatively straightforward; in other jurisdictions that have implemented or considered implementing an incentive scheme, STA profitability has declined during the pandemic—often quite substantially—and as a result the size of the necessary incentive is smaller now than it would have been before the pandemic. In Prince Edward County, by contrast, STA prices have skyrocketed in the last 18 months, and so the necessary financial incentives have grown correspondingly much larger.

```{r fig-4-3, include = TRUE, fig.cap = '(ref:fig-4-3)', fig.align = "center"}

figure_4_3_fun()

```
(ref:fig-4-3) *Average monthly subsidy required for transitioning an STA onto the long-term rental market*


### Restrictions on annual reservations

Many municipalities have set annual limits on per-property STA reservations as a way to limit STA activities to more casual, home sharing operations. In theory this is a reasonable idea, but in practice the enforcement burden makes reservation limits nearly impossible to implement well. To prove that a given STA has exceeded an annual reservation limit of, e.g., 60 nights, it would be necessary for enforcement officers to make 61 visits. Moreover, in a strongly seasonal destination such as Prince Edward County, it could still be economically viable to operate an STA for only a relatively small portion of the year and hold the property otherwise vacant. So even if it were possible to enforce a limit on reservations, this would be unlikely to achieve the goal of shifting STA activity to home sharing.


## Conclusions

This chapter has presented two models for STA licensing—a "Redistribute" model that entails redirecting STA activity away from dedicated commercial operations and towards home sharing instead, and a "Restrict" model that limits the number of STAs which are allowed to operate in order to limit the number of homes which are converted into STAs—alongside a set of other recommendations for regulatory reform. Our own judgement is that the "Redistribute" model applied County-wide provides the best balance of developing the positive impacts of STAs while mitigating the negative impacts, but each of the models represents a different set of trade-offs about which reasonable opinions may differ.

To conclude, however, we wish to reiterate an important constraint facing Prince Edward County in any attempt to regulate STAs, which is the dominance of second-home sales in the current housing market. Our analysis has indicating the ways in which STAs have contributed to a worsening housing affordability context for the County, but it has equally indicated that the single most important explanation for the pandemic-era explosion in housing prices and rents has been a second-home induced demand shock. Since the simplest substitute for a full-time STA is a second home, an implication of this fact is that both the "Redistribute" and "Restrict" models will likely clear room in the housing market which will be occupied primarily by an increase in second-home purchases. There is unfortunately no magic bullet capable of addressing this problem; although, as we noted above, one promising option is building more housing in typologies which are well-suited to local residents but less well-suited to second homes.

STA regulation should thus be understood as one piece of a larger system of necessary regulatory reform which would preserve and expand residents' access to affordable housing options in Prince Edward County. 

\newpage
