# Registration analysis

<style type="text/css">
  body{
  font-family: Futura, Helvetica, Arial;
}
</style>

<br>

```{r setup, include = FALSE, echo = FALSE}

library(here)
source(here("R", "01_startup.R"))

knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(include = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)

# Function for checking file modifications
mtime <- function(files) lapply(Sys.glob(files), function(x) file.info(x)$mtime)

# Set this to TRUE to output PDF figures
build_figures <- FALSE

qload(here("output", "/geometry.qsm"))
qload(here("output", "/str_processed.qsm"))
STA <- qread(here("output", "/STA.qs"))


```


``` {r objects, include = T}

library(kableExtra)

start_reg <- as.Date("2019-11-04")
end_reg <- as.Date("2020-09-04")
start_mora <- as.Date("2020-09-05")
end_mora <- as.Date("2021-06-22")

# Get properties active during the period when the registration was into effect
reg_property <- 
  property %>% 
  filter(active >= start_reg) %>% 
  filter(created <= end_reg) %>% 
  st_transform(32618)

n_reg_property <- 
  reg_property %>% 
  nrow() %>% 
  prettyNum(",")

total_licenses_all <- 
  STA %>% 
  st_drop_geometry() %>% 
  nrow()

valid_STA <-
  STA %>% 
  st_drop_geometry() %>% 
  filter(!is.na(expiry)) %>% 
  filter(status== "Active" | status == "Expired") %>% 
  nrow()

n_valid_STA_bc <- 
  STA %>% 
  st_drop_geometry() %>% 
  filter(!is.na(expiry)) %>% 
  filter(status== "Active" | status == "Expired") %>% 
  filter(created <= end_reg) %>% 
  nrow()

# Number of licenses per ward 
valid_STA_bc <- 
  STA %>% 
  filter(!is.na(expiry)) %>% 
  filter(status== "Active" | status == "Expired") %>% 
  filter(created <= end_reg)


# Find property_IDs within the buffer of each license
ids <- 
  valid_STA_bc %>% 
  st_buffer(dist = 200) %>% 
  st_intersection(.,reg_property) %>% 
  st_drop_geometry() %>% 
  group_by(id) %>% 
  summarize(property_IDs = list(property_ID))

ids_sta_pid <- 
  ids %>% 
  nrow()

unique_PIDs <- 
  valid_STA_bc %>% 
  st_buffer(dist = 200) %>% 
  st_intersection(.,reg_property) %>% 
  st_drop_geometry() %>% 
  distinct(property_ID, .keep_all = TRUE) %>% 
  nrow()

# Number of license application by ward
picton_n_license <- 
  valid_STA_bc %>% 
  st_join(EW, .) %>% 
  st_drop_geometry() %>% 
  count(ward) %>% 
  arrange(desc(n)) %>% 
  slice(1) %>% 
  pull(n)

hallo_n_license <- 
  valid_STA_bc %>% 
  st_join(EW, .) %>% 
  st_drop_geometry() %>% 
  count(ward) %>% 
  arrange(desc(n)) %>% 
  slice(2) %>% 
  pull(n)

well_n_license <- 
  valid_STA_bc %>% 
  st_join(EW, .) %>% 
  st_drop_geometry() %>% 
  count(ward) %>% 
  arrange(desc(n)) %>% 
  slice(3) %>% 
  pull(n)

total_n_license <- 
  valid_STA_bc %>% 
  st_drop_geometry() %>% 
  nrow()


```

``` {r tab-5-1, include = TRUE}

library(kableExtra)

STA_bc %>% 
  st_join(EW, .) %>% 
  count(ward) %>% 
  st_drop_geometry() %>% 
  add_row(ward = "Prince Edward County",
          n = n_valid_STA_bc) %>% 
  arrange(desc(n)) %>% 
  rename(Ward = ward,
         `Number of STA licenses emitted` = n) %>% 
  kbl(caption = paste0("Number of active STA licenses by ward, 4 Nov. 2019 to 4 Sept. 2021"),
      align = "lrrrr") %>%
  kable_styling(latex_options = "scale_down") %>% 
  row_spec(1, background = paste0(col_palette[2], "50"))

  
```

On 9 October 2018, Prince Edward County's city council approved a by-law allowing the registration of short-term accommodations (STA) on its territory. The new licensing program requires STA hosts to acquire one license per accommodation, to be renewed annually. The license fees depend on the number of guest room within the accommodation. The density of STAs operating within the county is set for each zone or area of the County, based on the zoning by-law. The percentage of STAs out of total existing dwelling units is set at 15%, with varying radius thresholds depending on the zones. The County recognizes four different types of STAs: traditional B&Bs, where owner lives on site; partial or entire principal residence STAs; partial or entire secondary residence STAs; and on-farm tourist homes. The licensing system came into effect in November 2019. The Covid-related travel restrictions started in the month of March 2020, with a moratorium on STA licensing established by the County on 30 September 2020. The STA registration was reinstated in the County for traditional B&Bs and principal residence STA on 22 June 2021, leaving other types of STAs accommodation still enable to be granted a license. The following chapter provides an analysis of Prince Edward County's STA licensing system in relation to short-term rentals.

# Overview of STA licenses

Since the coming into effect of the licensing system, a total of `r total_licenses_all` license applications were received. Out of the total applications, `r valid_STA` were approved, with `r n_valid_STA_bc` of them occurring prior to the moratorium on STA licenses. The ward that received the most applications was Picton with `r picton_n_license`, followed by Hallowell with `r hallo_n_license` license applications, and Wellington with `r well_n_license` (Table \@ref(tab:tab-5-1)). `r ids_sta_pid` of the `r valid_STA_bc` short-term accommodations with valid licenses had one or more STR property within a 200-meter buffer of their address (Airbnb and VRBO obfuscate listing locations by shifting them randomly up to 200 m). A total of `r unique_PIDs` active during the period of operation of the licensing system were located within the `r ids_sta_pid` STA buffers (a total of `r n_reg_property` properties were active on STR platforms during the same time period). The small proportion of STRs on platforms such 

```{r STA_portrait, include = T}

# Entire apt vs not entire apt
n_entire_apt <- 
  valid_STA_bc %>% 
  st_drop_geometry() %>% 
  filter(!is.na(entire_apt)) %>% 
  count(entire_apt) %>% 
  filter(entire_apt == TRUE) %>% 
  pull(n)

n_not_entire_apt <- 
  valid_STA_bc %>% 
  st_drop_geometry() %>% 
  filter(!is.na(entire_apt)) %>% 
  count(entire_apt) %>% 
  filter(entire_apt == FALSE) %>% 
  pull(n)

n_created_reg <- 
  property %>% 
  st_drop_geometry() %>% 
  filter(housing == TRUE) %>% 
  filter(created >= start_reg, created <= end_mora) %>% 
  group_by(created >= start_mora) %>% 
  summarize(n = n()) %>% 
  set_names(c("during_mora", "n")) %>% 
  filter(during_mora == FALSE) %>% 
  pull(n)

n_created_mora <- 
  property %>% 
  st_drop_geometry() %>% 
  filter(housing == TRUE) %>% 
  filter(created >= start_reg, created <= end_mora) %>% 
  group_by(created >= start_mora) %>% 
  summarize(n = n()) %>% 
  set_names(c("during_mora", "n")) %>% 
  filter(during_mora == TRUE) %>% 
  pull(n)

n_created_post_mora <- 
  property %>% 
  st_drop_geometry() %>% 
  filter(housing == TRUE) %>% 
  filter(created >= end_mora) %>% 
  nrow()

```

Out of the `r n_valid_STA_bc` STAs operating with a license during the 10 months of operation, `r n_entire_apt` licenses were for entire home properties, or housing units entirely operated at a STA. The larger proportion of accommodations (`r n_not_entire_apt`) were not entire housing units, but rather a portion of a property or a traditional short-term accommodation, such as a hotel or a camp site. One third (`r n_3_bed`) of the licenses were emitted for accommodations with 3 bedrooms. Close to a quarter (`r n_2_bed`) were for 2-bedroom accommodations and more than one sixth (`r n_4_bed`) were for 4-bedroom accommodations. 

# The effect of the STA registration system on the STR market

```{r make_fig_5_1}

figure_5_1_fun <- function(regular = "", condensed = "") {

property %>% 
  st_drop_geometry() %>% 
  count(created) %>% 
  mutate(n = slide_dbl(n, mean, .before = 6)) %>%
  filter(created >= start_2019) %>% 
  ggplot(aes(created, n)) +
  geom_line(color = col_palette[3]) +
  geom_text(aes(x = start_reg, y = 10,
                label = "STA registration\nin effect"),
            family = "Futura Condensed", 
            inherit.aes = FALSE, hjust = 1) +
  geom_text(aes(x = end_mora, y = 10,
                label = "STA registration\nmoratorium"),
            family = "Futura Condensed", 
            inherit.aes = FALSE, hjust = 0) +
  geom_segment(aes(x = as.Date("2019-10-01"), xend = as.Date("2019-12-01"),
                   y = 9, yend = 8),
               colour = "black",
               arrow = arrow(length = unit(0.1, "cm"),type = "open"))+
  geom_segment(aes(x = as.Date("2021-07-01"), xend = as.Date("2021-05-01"),
                   y = 9, yend = 8),
               colour = "black",
               arrow = arrow(length = unit(0.1, "cm"),type = "open"))+
  annotate("rect", xmin = start_reg, xmax = end_reg, fill = col_palette[1],
           ymin = -Inf, ymax = Inf, alpha = 0.2) +
  annotate("rect", xmin = start_mora, xmax = end_mora, fill = col_palette[2],
           ymin = -Inf, ymax = Inf, alpha = 0.2) +
  scale_x_date(name = NULL) +
  scale_y_continuous(name = NULL) +
  theme_minimal()
  
}

if (build_figures) {
  ggsave(here("output", "figures", "figure_5_1.pdf"), 
         plot = figure_5_1_fun("Futura", "Futura Condensed"), 
         width = 8, height = 4.2, units = "in", useDingbats = FALSE)
  
  extrafont::embed_fonts(here("output", "figures", "figure_5_1.pdf"))
}

```

The STA registration system operated from November 2019 to September 2020, up until Prince Edward County's city council voted on a STA registration moratorium amidst the Covid-19 pandemic (Countylive 2020). No new short-term rental accommodations were technically allowed to be created in Prince Edward County from the end of September 2020. The STA registration moratorium was lifted in the end of June 2021 for bed & breakfasts and owner-occupied STRs only (Countylive 2021). Looking at Figure \@ref(fig:fig-5-1), we can see the number of new properties created on Airbnb and/or VRBO per day, both during the STA registration period and the STA registration moratorium. The number of listings being created started to decrease at the end of 2019. During the 10-month STA registration period, a total of `r n_created_reg` new STR listings operating out of housing units were created. In the following 7-month moratorium, `r n_created_mora` new listings were created, with an additional `r n_created_post_mora` in the partial partial moratorium period, spanning from June 2021 to the end of August (the end of the available data). Both the registration and the moratorium did not seem to have a direct effect on the creation of new STRs, which has remained similar in both periods studied. However, we can see that a resurgence of new listings being created occurred during the 2021 peak season.

``` {r fig-5-1, include = TRUE, fig.cap = '(ref:fig-5-1)', fig.align = "center"}

figure_5_1_fun()

```

(ref:fig-5-1) _Number of new STR properties created by day, 7-day rolling average_

# How is the STA market related to Prince Edward County's housing market?

The STA registration database enabled the analysis of the license applications with the house sales that occurred from 2016 to today. This acts a an indicator of the contribution of short-term rental to the County's current housing situation. Have properties been sold to then apply to become a STA? Or conversingly, have properties been converted into STAs, and then sold on the market as short-term rental investment properties? 

```{r, include = T}

home_sales <- qread(here("output", "home_sales.qs"),
                              nthreads = availableCores())

sold_before_created <- 
home_sales %>% 
  st_drop_geometry() %>% 
  left_join(STA, by = "address") %>% 
  filter(!is.na(id)) %>%
  filter(sale_date <= created) %>% 
  count(address) %>% 
  nrow()

per_sold_before_created <- 
{sold_before_created / nrow(STA)} %>% 
  scales::percent(0.1)

sold_before_created_2017 <- 
home_sales %>% 
  st_drop_geometry() %>% 
  left_join(STA, by = "address") %>% 
  filter(!is.na(id)) %>%
  filter(sale_date <= created,
         sale_date >= "2017-01-01") %>% 
  count(address) %>% 
  nrow()

per_sold_before_created_2017 <- 
{sold_before_created_2017 / nrow(STA)} %>% 
  scales::percent(0.1)

sold_before_created_2018 <- 
home_sales %>% 
  st_drop_geometry() %>% 
  left_join(STA, by = "address") %>% 
  filter(!is.na(id)) %>%
  filter(sale_date <= created,
         sale_date >= "2018-01-01", sale_date <= "2018-12-31") %>% 
  count(address) %>% 
  nrow()

per_sold_before_created_2018 <- 
{sold_before_created_2018 / nrow(STA)} %>% 
  scales::percent(0.1)

sold_before_created_2019 <- 
home_sales %>% 
  st_drop_geometry() %>% 
  left_join(STA, by = "address") %>% 
  filter(!is.na(id)) %>%
  filter(sale_date <= created,
         sale_date >= "2019-01-01", sale_date <= "2019-12-31") %>% 
  count(address) %>% 
  nrow()

per_sold_before_created_2019 <- 
{sold_before_created_2019 / nrow(STA)} %>% 
  scales::percent(0.1)

sold_before_created_2020 <- 
home_sales %>% 
  st_drop_geometry() %>% 
  left_join(STA, by = "address") %>% 
  filter(!is.na(id)) %>%
  filter(sale_date <= created,
         sale_date >= "2020-01-01", sale_date <= "2020-12-31") %>% 
  count(address) %>% 
  nrow()

per_sold_before_created_2020 <- 
{sold_before_created_2020 / nrow(STA)} %>% 
  scales::percent(0.1)

created_before_sale <- 
home_sales %>% 
  st_drop_geometry() %>% 
  left_join(STA, by = "address") %>% 
  filter(!is.na(id)) %>%
  filter(sale_date >= created) %>% 
  count(address) %>% 
  nrow()

```

There are `r sold_before_created` housing units that were sold since 2016 that then applied for a STA license during the application period. Out of all the `r total_licenses_all` license applications received by the county, about a third (`r per_sold_before_created`) of them were for housing units that were bought after January 2016. `r per_sold_before_created_2017` license applications were for properties bought in 2017, `r per_sold_before_created_2018` for properties bought in 2018, `r per_sold_before_created_2019`in 2019, and finally `r per_sold_before_created_2020` in 2020. Registrations being linked to newly bought housing units shows that a number of units are bought with the purpose of turning them in STR, and the registration system allows for it. There was also a lower number of STA license applications (`r created_before_sale`) that were received for properties that were then sold shortly after, most likely as an investment property operating as a short-term rental.





